Inference methods comparison for the simplified Naomi model using Pareto-smoothed importance sampling

### 2023-05-23 src/docs_paper/paper.Rmd
Rename posterior shrinkage to posterior contraction and start move into manuscript #45

These processes are informed by household survey data.
For $\theta \in \{\rho, \alpha, \kappa\}$ let
$$
\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
$$
be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
The observed number of outcomes are $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
$$
m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
$$
is the Kish effective sample size [@kish1965survey].
We use a binomial working likelihood
$$
y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})
$$
to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following weighted aggregates
\begin{equation*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \quad
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \quad
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{equation*}

@@

Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
Then the Laplace approximation to $p(\btheta, \y)$ is given by
$$
\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
$$

@@

A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by the following weighted sum
$$
\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z),
$$
where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, with weights chosen so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
In adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate] the nodes and weights are shifted and rotated to suit the particular integrand, based on the mode, curvature at the mode, and a matrix decomposition of the inverse curvature.
Two possibilities for this matrix decomposition are the Cholesky decomposition and the spectral decomposition [@jackel2005note].

@@

We performed posterior predictive checks to assess the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.

@@

Let $\{\phi_i\}_{i = 1}^n$ be posterior marginal samples from some quantity with empirical cumulative distribution (ECDF) function $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is given by the maximum absolute difference between two ECDFs.
For each method we compare the KS statistics
\begin{equation*}
D_\bullet = \sup_\vartheta | F_\text{NUTS}(\vartheta) - F_\bullet(\vartheta)|.
\end{equation*}
See a summary of the results in Table and Figure \ref{fig:ks}, and full results available in the appendix.

@@

As marginal distributions, we are interested in assessing the accuracy of joint distributions.
Let $\{\bphi_i\}_{i = 1}^n$ be samples from a joint posterior.
Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios $r_s = p(\theta_s, y) / q(\theta)$ used in importance sampling.
Let
$$
p(y \, | \, \mu, \sigma, k) =
\begin{cases}
\frac{1}{\sigma} \left( 1 + k \left( \frac{y - \mu}{\sigma} \right) \right)^{- \frac{1}{k} - 1}, \quad k \neq 0 \\
\frac{1}{\sigma} \exp \left( \frac{y - \mu}{\sigma} \right), \quad k = 0
\end{cases}
$$
be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.

@@

Another way to compare joint distributions is via the maximum mean discrepancy [MMD; @gretton2006kernel].
Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of samples from a joint posterior obtained using different inference methods.
The MMD can be empirically estimated by
$$
\text{MMD}(\Phi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bphi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
$$
See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.

>>>>

These processes are informed by household survey data.
For $\theta \in \{\rho, \alpha, \kappa\}$ let
\begin{equation*}
\hat \theta_\mathcal{I} = \frac{\sum_j w_j \cdot\theta_j}{\sum_j w_j}
\end{equation*}
be weighted, aggregate survey observations, where $j$ indexes individuals across all strata $i \in \mathcal{I}$ and $w_j$ are design weights.
The observed number of outcomes are $y^{\hat \theta}_{\mathcal{I}} = m^{\hat \theta}_{\mathcal{I}} \cdot \hat \theta_{\mathcal{I}}$ where
\begin{equation*}
m^{\hat \theta}_\mathcal{I} = \frac{\left(\sum_j w_j\right)^2}{\sum_j w_j^2},
\end{equation*}
is the Kish effective sample size [@kish1965survey].
We use a binomial working likelihood
\begin{equation*}
y^{\hat \theta}_{\mathcal{I}} \sim \text{xBin}(m^{\hat \theta}_{\mathcal{I}}, \theta_{\mathcal{I}})
\end{equation*}
to model these aggregate observations, where $\theta_{\mathcal{I}}$ are the following weighted aggregates
\begin{equation*}
\rho_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i}{\sum_{i \in \mathcal{I}} N_i}, \quad
\alpha_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \alpha_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}, \quad
\kappa_{\mathcal{I}} = \frac{\sum_{i \in \mathcal{I}} N_i \rho_i \kappa_i}{\sum_{i \in \mathcal{I}} N_i \rho_i}.
\end{equation*}

@@

Let $\tilde p_\texttt{G}(\x \, | \, \btheta, \y) = \mathcal{N}(\x \, | \, \hat \x(\btheta), \mathbf{H}(\btheta)^{-1})$ be a Gaussian approximation to $p(\x \, | \, \btheta, \y)$ with mode and precision matrix given by
\begin{align*}
\hat \x(\btheta) &= \argmax_\x \log p(\y, \x, \btheta), \\
\mathbf{H}(\btheta) &= - \frac{\partial^2}{\partial x \partial x^\top} \log p(\y, \x, \btheta) \rvert_{\x = \hat \x(\btheta)}.
\end{align*}
Then the Laplace approximation to $p(\btheta, \y)$ is given by
\begin{equation*}
\tilde p_\texttt{LA}(\btheta, \y) = \frac{p(\y, \x, \btheta)}{\tilde p_\texttt{G}(\x \, | \, \btheta, \y)} \Big\rvert_{\x = \hat \x(\btheta)}
\end{equation*}

@@

A quadrature rule can be used to approximate the integral of a function $f: \mathcal{Z} \to \mathbb{R}$ by the following weighted sum
\begin{equation*}
\int_{\z \in \mathcal{Z}} f(\z) \text{d}\z \approx \sum_{\z \in \mathcal{Q}} f(\z) w(\z),
\end{equation*}
where $\z \in \mathcal{Q} \subset \mathcal{Z}$ are a set of nodes and $\omega: \mathcal{Q} \to \mathbb{R}$ is a weighting function.
In Gauss-Hermite quadrature [@davis1975methods] the nodes are selected as zeros of the Hermite polynomials, with weights chosen so as to interpolate functions of the form $f(\z) = \tilde f(\z) \exp(-\z^2)$.
In adaptive Gauss-Hermite quadrature [@naylor1982applications; @tierney1986accurate] the nodes and weights are shifted and rotated to suit the particular integrand, based on the mode, curvature at the mode, and a matrix decomposition of the inverse curvature.
Two possibilities for this matrix decomposition are the Cholesky decomposition and the spectral decomposition [@jackel2005note].

@@

### Posterior contraction

To assess the informativeness of the data on each model parameter $\phi$ we compared (Figure \ref{fig:contraction}) the prior variance $\sigma_\text{prior}^2(\phi)$ to the posterior variance $\sigma_\text{posterior}^2(\phi)$ via the posterior contraction [@schad2021toward].
\begin{equation*}
c(\phi) = 1 - \frac{\sigma_\text{posterior}^2(\phi)}{\sigma_\text{prior}^2(\phi)}.
\end{equation*}

```{r contraction, fig.cap="Posterior contraction."}
knitr::include_graphics("depends/posterior-contraction.png")
```

### Coverage

We assessed the coverage of our estimates via the uniformity of the data within each posterior marginal distribution.
Let $\{\phi_i\}_{i = 1}^n$ be posterior marginal samples.

@@

The two-sample Kolmogorov-Smirnov (KS) test statistic [@smirnov1948table] is given by the maximum absolute difference between two empirical cumulative distribution (ECDF) functions $F(\varphi) = \frac{1}{n} \sum_{i = 1}^n \mathbb{I}_{\phi_i \leq \varphi}$.
For each method we compare the KS statistics
\begin{equation*}
D_\bullet = \sup_\varphi | F_\text{NUTS}(\varphi) - F_\bullet(\varphi)|.
\end{equation*}
See a summary of the results in Table and Figure \ref{fig:ks}, and full results available in the appendix.

@@

As marginal distributions, we are interested in assessing the accuracy of joint distributions.
Let $\{\bphi_i\}_{i = 1}^n$ be samples from a joint posterior.
Pareto-smoothed importance sampling (PSIS) [@vehtari2015pareto, @yao2018yes] is a method for stabilising the ratios $r_s = p(\theta_s, y) / q(\theta)$ used in importance sampling.
Let
\begin{equation*}
p(y \, | \, \mu, \sigma, k) =
\begin{cases}
\frac{1}{\sigma} \left( 1 + k \left( \frac{y - \mu}{\sigma} \right) \right)^{- \frac{1}{k} - 1}, \quad k \neq 0 \\
\frac{1}{\sigma} \exp \left( \frac{y - \mu}{\sigma} \right), \quad k = 0
\end{cases}
\end{equation*}
be the generalised Pareto distribution with shape parameter $k$ and location-scale parameter $(\mu, \tau)$.
See a summary of the results in Table and Figure \ref{fig:psis}, and full results available in the appendix.

@@

Another way to compare joint distributions is via the maximum mean discrepancy [MMD; @gretton2006kernel].
Let $\Phi = \{\bphi_i\}_{i = 1}^n$ and $\Psi = \{\bpsi_i\}_{i = 1}^n$ be two sets of samples from a joint posterior obtained using different inference methods.
The MMD can be empirically estimated by
\begin{equation*}
\text{MMD}(\Phi, \Psi) = \sqrt{\frac{1}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bphi_j) - \frac{2}{n^2} \sum_{i, j = 1}^n k(\bphi_i, \bpsi_j) + \frac{1}{n^2} \sum_{i, j = 1}^n k(\bpsi_i, \bpsi_j)}.
\end{equation*}
See a summary of the results in Table and Figure \ref{fig:mmd}, and full results available in the appendix.

### 2023-05-23 src/naomi-simple_psis/psis.Rmd
Rename posterior shrinkage to posterior contraction and start move into manuscript #45

aghq_q_ll <- apply(mvn_ll, 1, FUN = function(row) matrixStats::logSumExp(log(aghq_weights) + row))
plot(aghq_q_ll)
```

@@

aghq_psis <- loo::psis(log_ratios = lw, r_eff = 1)
aghq_psis$diagnostics
```

>>>>

aghq_q_ll <- apply(mvn_ll, 1, FUN = function(row) matrixStats::logSumExp(log(aghq_lambda) + row))
plot(aghq_q_ll)
```

@@

aghq_psis <- loo::psis(log_ratios = r, r_eff = 1)
aghq_psis$diagnostics
```
